---
title: "Reproducible Research Project 1"
author: "Matt Jester"
date: "April 13, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F)
```

### About

This is the first assignment in the Reproducible Research class. The purpose is to answer a set of questions about a data set collected from an activity monitoring device such as a Fitbit, Nike Fuelband or Jawbone Up.

### Download and import research data
```{r}
temp <- tempfile()
download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip',temp)
StepsData <- read.delim(unzip(temp),sep = ',')
unlink(temp)
library(data.table)
```
  

### Questions
#### What is mean total number of steps taken per day?
>**Calculation of total steps per day and histogram of the resulting values.**
>```{r}
StepsbyDate <- aggregate(steps ~ date, StepsData, sum)
hist(StepsbyDate$steps, breaks = 30, main = 'Steps per Day', xlab = 'Number of Steps')
```
>
>**Determination of the mean and median steps taken per day.**
>```{r}
meansteps <- mean(StepsbyDate$steps)
mediansteps <- median(StepsbyDate$steps)
````
> The mean is ``r format(meansteps, nsmall = 2)`` and the median is ``r format(mediansteps, nsmall = 2)``.
  
#### What is the average daily activity pattern?
>**Determine average steps per 5 minute interval and plot the results.**
>```{r}
# Total steps per interval
StepsbyInterval <- aggregate(steps ~ interval, StepsData, sum)

# Days with values for each interval
CompleteIntervals <- StepsData[complete.cases(StepsData),]
CountsbyInterval <- as.data.frame(table(CompleteIntervals$interval))
colnames(CountsbyInterval) = c('interval', 'datecount')

# Merge step count totals with count of days
AvgStepsbyInterval <- merge(StepsbyInterval, CountsbyInterval)
AvgStepsbyInterval$AvgSteps <- with(AvgStepsbyInterval, steps / datecount)

# Plot the average steps per interval
plot(AvgStepsbyInterval$interval, AvgStepsbyInterval$AvgSteps, type = 'l',
     ylab = 'steps', xlab = 'interval', main = 'Average Steps per 5 min Interval')
```
>
>**Determination of the interval which averages the most steps.**
>```{r}
maxsteps <- StepsbyInterval[which.max(StepsbyInterval[,2]),1]
```
> The interval with the most steps on average is ``r maxsteps``.
  
  
#### Imputing missing values
> How many values are missing from the data set?
>```{r}
incomplete <- sum(!complete.cases(StepsData))
```
>
> There are ``r incomplete`` values missing from the data set.
>  
> We could assume that the average values by interval determined above would be acceptable for imputing each missing value. However, is there a variance based on day of the week which should be considered? By using the ```weekdays``` and ```mean``` functions on the totals by day, we can see if there are large differences for each day of the week. The able below summarizes that data as well a the number of missing values by day of the week.
>
>Day  | Avg Step Count| Missing Values  
---------- | ---------- | ---------- 
Sunday    | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Sunday',2]), nsmall = 2)``| ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Sunday',]))``        
Monday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Monday',2]), nsmall = 2)`` | ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Monday',]))``               
Tuesday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Tuesday',2]), nsmall = 2)``| ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Tuesday',]))``                
Wednesday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Wednesday',2]), nsmall = 2)``        | ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Wednesday',]))``        
Thursday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Thursday',2]), nsmall = 2)``        | ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Thursday',]))``        
Friday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Friday',2]), nsmall = 2)`` | ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Friday',]))``               
Saturday        | ``r format(mean(StepsbyDate[weekdays(as.Date(StepsbyDate$date))=='Saturday',2]), nsmall = 2)``        | ``r sum(!complete.cases(StepsData[weekdays(as.Date(StepsbyDate$date))=='Saturday',]))``        
>  
> Based on these values we should impute values based on day of the weeks as well as the interval.
>```{r}
StepsData <- data.table(StepsData)
AvgforImputing <- StepsData[,list(
        SunAvg = mean(steps[weekdays(as.Date(date))=='Sunday'], na.rm = T),
        MonAvg = mean(steps[weekdays(as.Date(date))=='Monday'], na.rm = T),
        TueAvg = mean(steps[weekdays(as.Date(date))=='Tuesday'], na.rm = T),
        WedAvg = mean(steps[weekdays(as.Date(date))=='Wednesday'], na.rm = T),
        ThuAvg = mean(steps[weekdays(as.Date(date))=='Thursday'], na.rm = T),
        FriAvg = mean(steps[weekdays(as.Date(date))=='Friday'], na.rm = T), 
        SatAvg = mean(steps[weekdays(as.Date(date))=='Saturday'], na.rm = T)), 
        by = c('interval')]
StepsData_i <- merge(StepsData, AvgforImputing, 'interval')
StepsData_i$steps <- with(StepsData_i, 
        ifelse(weekdays(as.Date(date))=='Sunday' & is.na(steps), SunAvg,
        ifelse(weekdays(as.Date(date))=='Monday' & is.na(steps), MonAvg,
        ifelse(weekdays(as.Date(date))=='Tuesday' & is.na(steps), TueAvg,
        ifelse(weekdays(as.Date(date))=='Wednesday' & is.na(steps), WedAvg,
        ifelse(weekdays(as.Date(date))=='Thursday' & is.na(steps), ThuAvg,
        ifelse(weekdays(as.Date(date))=='Friday' & is.na(steps), FriAvg,
        ifelse(weekdays(as.Date(date))=='Saturday' & is.na(steps), SatAvg, 
        steps))))))))
StepsData_i <- StepsData_i[,c(3,1:2)]
```
>  
> **How does the imputed data impact the distribution of steps by day as well as the mean and median values?**
>```{r}
StepsbyDate_i <- aggregate(steps ~ date, StepsData_i, sum)
hist(StepsbyDate_i$steps, breaks = 30, main = 'Steps per Day (w/imputed values)', xlab = 'Number of Steps')
```
>  
> The number of days with 9,000 & 12,000 steps have each increased by 1 while the number of days with 13,000 steps has increased by 4 to tie the number of days with 11,000 steps at 10 days each.
>
> **Determination of the new mean and median steps taken per day.**
>```{r}
meansteps_i <- mean(StepsbyDate_i$steps)
mediansteps_i <- median(StepsbyDate_i$steps)
```
> With the imputed missing values, the mean is ``r format(meansteps_i, nsmall = 2)`` and the median is ``r format(mediansteps_i)``. These values represent increases of ``r format(meansteps_i - meansteps, nsmall = 2)`` for the mean and ``r mediansteps_i - mediansteps`` for the median.
  
#### Are there differences in activity patterns between weekdays and weekends?
> A factor variable can be introduced to identify which days are weekends and which are weekdays.
>```{r}
WEDays <- c('Saturday', 'Sunday')
StepsData_i$DoW = as.factor(ifelse(is.element(weekdays(as.Date(StepsData_i$date))
        ,WEDays), "Weekend", "Weekday"))
StepsbyInterval_i <- aggregate(steps ~ interval + DoW, StepsData_i, mean)
```
>  
> With the factor variable in place, we can now graph the data to compare weekdays to weekend.
>```{r}
library(lattice)
xyplot(StepsbyInterval_i$steps ~ StepsbyInterval_i$interval | StepsbyInterval_i$DoW, 
       main="Average Steps per Day by Interval", xlab = "Interval", ylab = "Steps", 
       layout = c(1, 2), type = "l")
```
>  
> Utilizing this pair of plots, we see that while the peak is higher on week days, the sustained average number of steps for each interval is higher on weekends.
